Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-09-04 jso.qmd
Update jso.qmd
working on Practicalities section

  The projection pursuit (PP) guided tour interactively optimises a criteria function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimisation of some PP indexes can be non-trivial, if they are non-smooth functions, or the optima has a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, methods for calculating the smoothness and squintability properties of the PP index are proposed. There are used to assess the optimiser performance in the presence of PP index complexities. A simulation study illustrates the use of these performance metrics to compare the JSO with existing optimisation methods available for the guided tour.  The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability are available in the `ferrn` package. 

@@

# Simulation details {#sec-sim-deets}

@@

# Discussion {#sec-discussion}

This section discusses the insights in implementation yielded from the previous sections to guide the usage of projection pursuit guided tour amongst practitioners. Smoothness and squintability in a projection pursuit problem can now be measured given the target shape, the chosen index function, and the data dimension. In practice, these metrics can be used to characterise the difficulty of the optimisation task and inform the choice of optimiser. For instance, a large squintability value indicates a distinct difference in index values between optimal regions and others. When one jellyfish enters the optimal region, the high index value it generates will be clearly distinguishable from spurious values due to this pronounced difference and lead other jellyfish to move toward the optimal region. In contrast, a smaller squintability value results in a less distinct separation between desirable and undesirable regions, making it more difficult for other jellyfish to accurately navigate toward the target area. In addition, a larger smoothness value implies a smoother index function, which suggests the use of both gradient-based and random sampling optimisers.

@@

The new JSO is integrated with the current implementation of the projection pursuit guided tour in the `tourr` package, and can be further examined using PP optimisation diagnostics in the `ferrn` package. The recommended approach is to conduct the optimisation off-line, extract the bases of a selected jellyfish, and then use the planned tour to follow selected jellyfish. These tools are also available in the `tourr` package.

@@

  - Smoothness and squintability can be calculated and compared across different problems to understand their relative complexity. Once a projection pursuit problem is fully characterised (by the shape-to-find, the index function used, and the data dimension), increasing JSO hyper-parameters can enhance the search effectiveness, however, the resulting increase in computational complexity will also need to be considered. - modelling
-->
```
# Acknowledgement

>>>>

  The projection pursuit (PP) guided tour interactively optimises a criteria function known as the PP index, to explore high-dimensional data by revealing interesting projections. Optimisation of some PP indexes can be non-trivial, if they are non-smooth functions, or the optima has a small "squint angle", detectable only from close proximity. To address these challenges, this study investigates the performance of a recently introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for optimising PP indexes. The performance of JSO for visualising data is evaluated across various hyper-parameter settings and compared with existing optimisers. Additionally, methods for calculating the smoothness and squintability properties of the PP index are proposed. They are used to assess the optimiser performance in the presence of PP index complexities. A simulation study illustrates the use of these performance metrics to compare the JSO with existing optimisation methods available for the guided tour.  The JSO algorithm has been implemented in the R package, `tourr`, and functions to calculate smoothness and squintability are available in the `ferrn` package. 

@@

# Simulation study {#sec-sim-deets}

@@

# Practicalities {#sec-discussion}

## Computing the metrics for your new index

## Using the JSO in a PPGT

```{r}
#| eval: false
#| echo: false
jso_optima <- save_history(flea[,1:6], 
                           guided_tour(holes()), 
                           search_f = search_jellyfish)
```
<!--
This section discusses the insights in implementation yielded from the previous sections to guide the usage of projection pursuit guided tour amongst practitioners. Smoothness and squintability in a projection pursuit problem can now be measured given the target shape, the chosen index function, and the data dimension. In practice, these metrics can be used to characterise the difficulty of the optimisation task and inform the choice of optimiser. For instance, a large squintability value indicates a distinct difference in index values between optimal regions and others. When one jellyfish enters the optimal region, the high index value it generates will be clearly distinguishable from spurious values due to this pronounced difference and lead other jellyfish to move toward the optimal region. In contrast, a smaller squintability value results in a less distinct separation between desirable and undesirable regions, making it more difficult for other jellyfish to accurately navigate toward the target area. In addition, a larger smoothness value implies a smoother index function, which suggests the use of both gradient-based and random sampling optimisers.

@@

The new JSO is integrated with the current implementation of the projection pursuit guided tour in the `tourr` package, and can be further examined using PP optimisation diagnostics in the `ferrn` package. The recommended approach is to conduct the optimisation off-line, extract the bases of a selected jellyfish, and then use the planned tour to follow selected jellyfish. These tools are also available in the `tourr` package.
-->

@@

  - Smoothness and squintability can be calculated and compared across different problems to understand their relative complexity. Once a projection pursuit problem is fully characterised (by the shape-to-find, the index function used, and the data dimension), increasing JSO hyper-parameters can enhance the search effectiveness, however, the resulting increase in computational complexity will also need to be considered. - modelling
-->
```

# Acknowledgement

### 2024-09-04 jso.tex
working on Practicalities section

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optima has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. There are used to assess the optimiser performance
in the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}

@@

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, left=2mm, rightrule=.15mm, opacityback=0, title={Algorithm: Jellyfish Optimizer Pseudo Code}, breakable, opacitybacktitle=0.6, colback=white, colframe=quarto-callout-note-color-frame, toprule=.15mm, toptitle=1mm, arc=.35mm, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black]

@@

\section{Simulation details}\label{sec-sim-deets}

@@

\section{Discussion}\label{sec-discussion}

This section discusses the insights in implementation yielded from the
previous sections to guide the usage of projection pursuit guided tour
amongst practitioners. Smoothness and squintability in a projection
pursuit problem can now be measured given the target shape, the chosen
index function, and the data dimension. In practice, these metrics can
be used to characterise the difficulty of the optimisation task and
inform the choice of optimiser. For instance, a large squintability
value indicates a distinct difference in index values between optimal
regions and others. When one jellyfish enters the optimal region, the
high index value it generates will be clearly distinguishable from
spurious values due to this pronounced difference and lead other
jellyfish to move toward the optimal region. In contrast, a smaller
squintability value results in a less distinct separation between
desirable and undesirable regions, making it more difficult for other
jellyfish to accurately navigate toward the target area. In addition, a
larger smoothness value implies a smoother index function, which
suggests the use of both gradient-based and random sampling optimisers.

This study enables comparison of new indices with existing indices with
the proposed smoothness and squintability measures, as well as
benchmarking with existing well-known data sets. Note that when
comparing between multiple optimisers, the smoothness and squintability
metrics proposed in this paper are relative. Comparison should be made
with common parameters, i.e.~the same seed for generating random bases
and the same binning width for squintability. Furthermore, the range of
the index values should be standardised to the {[}0, 1{]} interval as
demonstrated for the \texttt{holes} and \texttt{TIC} indexes.

As a population-based metaheuristic method, JSO requires evaluating
multiple jellyfish in each iteration, which increased its computational
complexity compared to single-candidate random sampling methods. This
additional computational costs should be taken into consideration when
choosing optimisers. Tuning JSO hyperparameters, such as the number of
jellyfish and the maximum number of attempts, can enhance the likelihood
of finding the optimal region at the expense of computational costs.
However, in some cases, increasing these parameters may not be effective
if the index surface is complex or has low squintability.

The new JSO is integrated with the current implementation of the
projection pursuit guided tour in the \texttt{tourr} package, and can be
further examined using PP optimisation diagnostics in the \texttt{ferrn}
package. The recommended approach is to conduct the optimisation
off-line, extract the bases of a selected jellyfish, and then use the
planned tour to follow selected jellyfish. These tools are also
available in the \texttt{tourr} package.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. Nicolas
Langrené. Email: \href{mailto:nicolaslangrene@uic.edu.cn}{\nolinkurl{nicolaslangrene@uic.edu.cn}}. Jessica
Wai Yin
Leung. Email: \href{mailto:Jessica.Leung@monash.edu}{\nolinkurl{Jessica.Leung@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Austin, United
States\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash
University, Melbourne, Australia\\ \textsuperscript{3} Department of
Mathematical Sciences, Guangdong Provincial Key Laboratory of
Interdisciplinary Research and Application for Data Science,
BNU-HKBU~United~International~College, Zhuhai, China
\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. Optimisation of some PP
indexes can be non-trivial, if they are non-smooth functions, or the
optima has a small ``squint angle'', detectable only from close
proximity. To address these challenges, this study investigates the
performance of a recently introduced swarm-based algorithm, Jellyfish
Search Optimiser (JSO), for optimising PP indexes. The performance of
JSO for visualising data is evaluated across various hyper-parameter
settings and compared with existing optimisers. Additionally, methods
for calculating the smoothness and squintability properties of the PP
index are proposed. They are used to assess the optimiser performance in
the presence of PP index complexities. A simulation study illustrates
the use of these performance metrics to compare the JSO with existing
optimisation methods available for the guided tour. The JSO algorithm
has been implemented in the R package, \texttt{tourr}, and functions to
calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
projection pursuit\sep jellyfish search optimiser
(JSO)\sep optimisation\sep grand tour\sep high-dimensional data\sep 
exploratory data analysis
\end{keywords}

@@

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, coltitle=black, bottomrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, title={Algorithm: Jellyfish Optimizer Pseudo Code}, left=2mm, opacitybacktitle=0.6, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white]

@@

\section{Simulation study}\label{sec-sim-deets}

@@

\section{Practicalities}\label{sec-discussion}

\subsection{Computing the metrics for your new
index}\label{computing-the-metrics-for-your-new-index}

\subsection{Using the JSO in a PPGT}\label{using-the-jso-in-a-ppgt}
